Gradual Soundness: Lessons from Static Python

### 2021-11-01 ecoop-2022/paper.tex
ecoop: first draft of theorems and types

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}

@@

@@

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\defeq}{=}
\newcommand{\mfeq}{=}
\newcommand{\langmid}{\mathrel{\mathbf{\Big\vert}}}
\newenvironment{langarray}{\(\def\arraystretch{1.5}\begin{array}{l@{\hspace{2mm}}c@{\hspace{2mm}}l}}{\end{array}\)}

\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{NoneType}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptfun}[2]{#1 \rightarrow #2}
\newcommand{\sptunion}[2]{#1 \cup #2}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptpydict}[2]{\paramtype{\typefont{PyDict}}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}

\newcommand{\spexpr}{e}
\newcommand{\spvalue}{v}

\newcommand{\sprred}{\rightarrow^*}

\newcommand{\mfapply}[2]{#1\,(#2)}
\newcommand{\mffont}[1]{\mathit{#1}}
\newcommand{\mftypeF}[1]{\mfapply{\mffont{F}}{#1}}
\newcommand{\mfopt}[1]{\mfapply{\mffont{opt}}{#1}}

\newcommand{\sperror}{\mathrm{Error}}

@@

\begin{figure}[t]
  %% ported types from redex model
  \begin{langarray}
    \sptype & \defeq &
      \sptdyn \langmid
      \sptnone \langmid
      \sptclass \langmid
      \sptinstanceof{\sptclass} \langmid
      \sptfun{\sptype}{\sptype} \langmid
      \sptunion{\sptclass}{\sptclass}
  \\
    \sptclass & \defeq &
      \sptint \langmid
      \sptfloat \langmid
      \sptobject \langmid
      \sptpydict{\sptclass}{\sptclass} \langmid
      \sptchkdict{\sptclass}{\sptclass} \langmid
      \mbox{(user defined class)}
  \end{langarray}

  \begin{center}\parbox{0.8\columnwidth}{
    Abbreviation: $\sptoptional{\sptclass} \defeq \sptunion{\sptclass}{\sptnone}$

    EXCEPT THAT $\sptoptional{\sptclass}$ may be a generic parameter just like any other class $\sptclass$

    Function types cannot expect or return other function types.
  }\end{center}

  \caption{Surface Types}
  \label{f:surface-types}
\end{figure}

@@

\begin{figure}[t]
  \(
    \mftypeF{\sptype_0}
    \mfeq
    \left\{\begin{array}{ll}
      \sptpydict{}{} & \mbox{if $\sptype_0 = \sptpydict{\sptclass}{\sptclass}$}
    \\
      \sptype_0 & \mbox{otherwise}
    \end{array}\right.
  \)

  \caption{Surface Types to Evaluation Types}
  \label{f:surface-to-eval-types}
\end{figure}

\subsection{Properties}

\begin{theorem}[Type Soundness]
  If\ \(\vdash \spexpr_0 : \sptype_0\)
  then one of the following holds:
  \begin{itemize}
    \item
      \(\spexpr_0 \sprred \spvalue_0
        \mbox{ and }
        \vdash \spvalue_0 : \mftypeF{\sptype_0}
      \)
    \item
      \(\spexpr_0\) diverges
    \item
      \(\spexpr_0 \sprred \sperror\)
  \end{itemize}
\end{theorem}

\begin{theorem}[Optimization Soundness]
  If\ \(\vdash \spexpr_0 : \mftypeF{\sptype_0}\)
  then\ \(\vdash \mfopt{\spexpr_0} : \mftypeF{\sptype_0}\)
\end{theorem}

\begin{theorem}[Optimization Effectiveness]
  FILL how do we know optimizations are useful rather than trivial?
  %% - length of eval(e) vs. eval(opt(e)) might be longer (good) or shorter (bad)
  %%   for general heuristic opts.
  %% - is [opt(e) != e] true for interesting programs?
\end{theorem}
